---
title: "Project 1"
output: html_notebook
---

Data set: regression 1 (fishermen - do not use the variable MeHg)

Group: Urvan Christen, Amandine Goffeney, Joseph Vermeil, Lucile Vigué

# Data description

Source: N.B. Al-Majed and M.R. Preston (2000). "Factors Influencing the Total
Mercury and Methyl Mercury in the Hair of Fishermen in Kuwait," 
Environmental Pollution, Vol. 109, pp. 239-250

Description: Factors related to mercury levels among fishermen and a control
group of non-fishermen.

Variables/names

- Fisherman indicator  (_fisherman_)
- Age in years  (_age_)
- Residence Time in years   (_restime_)
- Height in cm    (_height_)
- Weight in kg    (_weight_)
- Fish meals per week    (_fishmlwk_)
- Parts of fish consumed: 0=none, 1=muscle tissue only, 2=mt and sometimes whole fish, 3=whole fish  (_fishpart_)
- Methyl Mercury in mg/g    (_MeHg_)
- Total Mercury in mg/g     (_TotHg_)

# Imports and loading data

```{r}
# Load here the libraries and packages you need for the rest of the analysis.
```


```{r}
DATA_FILE = "fishermen_mercury.csv" # Location of the data csv file
dataset = read.csv(DATA_FILE) # Loading the data
dataset$MeHg <- NULL # We are asked not to use the variable MeHg
dataset
```


Let's separate the table into the table of interest (the fisherman) vs the control table (the non fisherman).
```{r}
dataset.fisherman=dataset[which(dataset$fisherman==1),]
dataset.non_fisherman=dataset[which(dataset$fisherman==0),]
```

We can now do some regression on both datasets and compare the results with some statistical tests? 

```{r}

```


