---
title: "Project 1"
output: html_notebook
---

Data set: regression 1 (fishermen - do not use the variable MeHg)

Group: Urvan Christen, Amandine Goffeney, Joseph Vermeil, Lucile Vigué

# Data description

Source: N.B. Al-Majed and M.R. Preston (2000). "Factors Influencing the Total
Mercury and Methyl Mercury in the Hair of Fishermen in Kuwait," 
Environmental Pollution, Vol. 109, pp. 239-250

Description: Factors related to mercury levels among fishermen and a control
group of non-fishermen.

Variables/names

- Fisherman indicator  (_fisherman_)
- Age in years  (_age_)
- Residence Time in years   (_restime_)
- Height in cm    (_height_)
- Weight in kg    (_weight_)
- Fish meals per week    (_fishmlwk_)
- Parts of fish consumed: 0=none, 1=muscle tissue only, 2=mt and sometimes whole fish, 3=whole fish  (_fishpart_)
- Methyl Mercury in mg/g    (_MeHg_)
- Total Mercury in mg/g     (_TotHg_)

# Proposition of analysis 

First : we should maybe show that the study makes sense, by showing that the mean of TotHg is significantly different for the fisherman and the control group.

Then, we could look for possible outliers, that we will eventually take out of the study afterwards.

Then, we could fit 2 or 3 models (with or without interaction terms, with more or less degrees of interaction), and determine with anova comparison, or AIC, which one we will use to draw conclusions; eventually also plot some QQ and things to show that ok the model is acceptable and to maybe find other outliers.

Proposition of 3 models (they can of course be changed, it's just a propostion) : 
- TotHg ~ restime + height + weight + fishmlwk + fishpart 

- TotHg ~ (restime + height + weight + fishmlwk + fishpart)^2

At last, we could take the "best" model to draw conclusions about which variables explain significantly the concentration TotHg, and what are the coefficients. 

In all the calculations, we should (to my opinion) separate the fisherman and the control group (to find the best model and do calculations).

# Imports and loading data

```{r}
# Load here the libraries and packages you need for the rest of the analysis.
```


```{r}
DATA_FILE = "fishermen_mercury.csv" # Location of the data csv file
dataset = read.csv(DATA_FILE) # Loading the data
dataset$MeHg <- NULL # We are asked not to use the variable MeHg
dataset
```


Let's separate the table into the table of interest (the fisherman) vs the control table (the non fisherman).
```{r}
dataset.fisherman=dataset[which(dataset$fisherman==1),]
dataset.non_fisherman=dataset[which(dataset$fisherman==0),]
```

We can now do some regression on both datasets and compare the results with some statistical tests? 

```{r}

```


