---
title: 'Project 1: regression 1 (fishermen - do not use the variable MeHg)'
author: "Urvan Christen, Amandine Goffeney, Joseph Vermeil, Lucile Vigué"
date: "March 17, 2019"
output:
  pdf_document:
    df_print: paged
geometry: margin=2.5cm
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[LE,RO]{Christen, Goffeney, Vermeil, Vigué}
fontsize: 12pt
df_print: paged
bibliography: bibliography.bib
---


# Data description

Source: [@al2000factors]

Description: Factors related to mercury levels among fishermen and a control
group of non-fishermen.

Variables/names

- Fisherman indicator  (_fisherman_)
- Age in years  (_age_)
- Residence Time in years   (_restime_)
- Height in cm    (_height_)
- Weight in kg    (_weight_)
- Fish meals per week    (_fishmlwk_)
- Parts of fish consumed: 0=none, 1=muscle tissue only, 2=mt and sometimes whole fish, 3=whole fish  (_fishpart_)
- Methyl Mercury in mg/g    (_MeHg_)
- Total Mercury in mg/g     (_TotHg_)



# Imports and loading data

```{r, echo=FALSE, message=FALSE}
# Load here the libraries and packages you need for the rest of the analysis.
library(tidyverse)
library("MASS")
library(GGally)
library(car) # contains the vif() and leveneTest() functions
```


```{r, echo=FALSE, message=FALSE}
DATA_FILE = "fishermen_mercury.csv" # Location of the data csv file
dataset = read.csv(DATA_FILE) # Loading the data
dataset$MeHg <- NULL # We are asked not to use the variable MeHg
dataset$LogTotHg <- log(dataset$TotHg)
dataset$logFishmlwk <- log(dataset$fishmlwk + 1)
dataset
```
We have the following continuous variables:
- Age in years  (_age_)
- Residence Time in years   (_restime_)
- Height in cm    (_height_)
- Weight in kg    (_weight_)
- Fish meals per week    (_fishmlwk_)
- Total Mercury in mg/g     (_TotHg_)

We have the following categorical or boolean variables:
- Fisherman indicator  (_fisherman_)
- Parts of fish consumed: 0=none, 1=muscle tissue only, 2=mt and sometimes whole fish, 3=whole fish  (_fishpart_)

NB: pour _restime_ et _fishmlwk_ je ne sais pas trop si on les compte comme catégoriques ou pas.

```{r, echo=FALSE, message=FALSE}
dataset$fisherman <- factor(dataset$fisherman)
#dataset$fishmlwk <- factor(dataset$fishmlwk)
dataset$fishpart <- factor(dataset$fishpart)
```


Let's separate the table into the table of interest (the fisherman) vs the control table (the non fisherman).
```{r, echo=FALSE, message=FALSE}
dataset.fisherman=dataset[which(dataset$fisherman==1),]
dataset.non_fisherman=dataset[which(dataset$fisherman==0),]
```
We can now do some regression on both datasets and compare the results with some statistical tests? 

# Exploratory analysis

## Summarize the dataset

```{r, echo=FALSE, message=FALSE}
summary(dataset)
```

## Plots
### Pairwise behaviour of explnatory variables

```{r, echo=FALSE, message=FALSE}
ggpairs(dataset[, c("fisherman", "age", "weight", "restime", "fishmlwk", "fishpart")])
ggpairs(dataset.fisherman[, c("fisherman", "age", "weight", "restime", "fishmlwk", "fishpart")])
ggpairs(dataset.non_fisherman[, c("fisherman", "age", "weight", "restime", "fishmlwk", "fishpart")])
```

### Plots of factored data
```{r, echo=FALSE, message=FALSE}
ggplot(dataset, aes(fisherman, TotHg)) + geom_boxplot()
ggplot(dataset, aes(age, TotHg, colour = fisherman)) + geom_point() #+ scale_y_log10() + scale_x_log10() ##Uncomment for log scales
ggplot(dataset, aes(restime, TotHg, colour = fisherman)) + geom_point()
ggplot(dataset, aes(height, TotHg, colour = fisherman)) + geom_point() #+ scale_y_log10() + scale_x_log10() 
ggplot(dataset, aes(weight, TotHg, colour = fisherman)) + geom_point() #+ scale_y_log10() + scale_x_log10()
ggplot(dataset, aes(fishmlwk, TotHg, colour = fisherman)) + geom_boxplot()
ggplot(dataset, aes(fishpart, TotHg, colour = fisherman)) + geom_boxplot()
```

## Analyze

### 1-D linear models

#### Age

##### Code

```{r, echo=FALSE, message=FALSE}
hg.form.age = TotHg ~ age # 1-D regression

hg.lm.age = lm(hg.form.age, data=dataset) # Whole population
hg.lmFi.age = lm(hg.form.age, data=dataset.fisherman) # Fishermen population
hg.lmNF.age = lm(hg.form.age, data=dataset.non_fisherman) # Control population

# Diagnostic plots
par(mfrow=c(2,2))
plot(hg.lm.age)
mtext("TotHg as a function of age - Whole population", side = 3, line = -2, outer = TRUE, cex = 1.5)

par(mfrow=c(2,2))
plot(hg.lmFi.age)
mtext("TotHg as a function of age - Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)


par(mfrow=c(2,2))
plot(hg.lmNF.age)
mtext("TotHg as a function of age - Non-Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)
```

##### Comments:

- Residuals seem all in all well distributed in all three populations suggesting that the distribution of TotHg as a function of age is homoscedastic.
- Normal Q-Q fits a line for Non-Fishermen population reinforcing the assumption that the distribution of TotHg given an age follows a normal distribution.
- Normal Q-Q suggest however a heavy right tail for Fishermen population on this same distribution

#####Test of homoskedasticity with a Breusch-Pagan test (Joseph)
- This is a test for homoskedasticity of the data. The null hypothesis is homoskedasticity, and ncvTest calculates a p-value.
```{r, echo=FALSE, message=FALSE}
ncvTest(hg.lm.age) # Output: Chisquare = 10.93327, Df = 1, p = 0.00094453
ncvTest(hg.lmFi.age) # Output: Chisquare = 4.561776, Df = 1, p = 0.032693
ncvTest(hg.lmNF.age) # Output: Chisquare = 0.2004155, Df = 1, p = 0.65439
```
These results suggest that only the non-Fisherman pop has homoskedastic HgTot vs. Age values.

#### Restime

##### Code

```{r, echo=FALSE, message=FALSE}
hg.form.restime = TotHg ~ restime

hg.lm.restime = lm(hg.form.restime, data=dataset)
hg.lmFi.restime = lm(hg.form.restime, data=dataset.fisherman)
hg.lmNF.restime = lm(hg.form.restime, data=dataset.non_fisherman)

par(mfrow=c(2,2))
plot(hg.lm.restime)
mtext("TotHg as a function of restime - Whole population", side = 3, line = -2, outer = TRUE, cex = 1.5)

par(mfrow=c(2,2))
plot(hg.lmFi.restime)
mtext("TotHg as a function of restime - Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)


par(mfrow=c(2,2))
plot(hg.lmNF.restime)
mtext("TotHg as a function of restime - Non-Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)
```

##### Comments

- There is a great difference of the distribution of *restime* between Fishermen and control population. In the control population it takes a complete range of values whereas in the control population, the distribution is discrete and takes only 3 (even 2) different values. Thus it might be difficult to draw conclusions on whether *restime* is correlated or not with the *TotHg* value. 

- There also is this problem in Fishermen population of the right long tail and the left short tail for *TotHg* distribution. **It might be useful to use a log scale.**

- Residuals in Fishermen population for fitted *TotHg* as a function of *restime* are all in all well distributed around 0 for all values of *restime* suggesting an homoscedasticity of the distribution of *TotHg* according to *restime*.

- However, there are some residuals with very high positive values when there are none with very "high" negative values, suggesting some possible bias in the distribution.

#####Test of homoskedasticity with a Breusch-Pagan test (Joseph)
- This is a test for homoskedasticity of the data. The null hypothesis is homoskedasticity, and ncvTest calculates a p-value.
```{r, echo=FALSE, message=FALSE}
ncvTest(hg.lm.restime ) # Output: Chisquare = 0.2252327, Df = 1, p = 0.63508
ncvTest(hg.lmFi.restime ) # Output: Chisquare = 0.9487475, Df = 1, p = 0.33004
ncvTest(hg.lmNF.restime ) # Output: Chisquare = 0.2388775, Df = 1, p = 0.62502
```
This time those results suggest that all 3 pop have homoskedastic HgTot vs. restime values.


#### Weight

##### Code

```{r, echo=FALSE, message=FALSE}
hg.form.weight = TotHg ~ weight

hg.lm.weight = lm(hg.form.weight, data=dataset)
hg.lmFi.weight = lm(hg.form.weight, data=dataset.fisherman)
hg.lmNF.weight = lm(hg.form.weight, data=dataset.non_fisherman)

par(mfrow=c(2,2))
plot(hg.lm.weight)
mtext("TotHg as a function of weight - Whole population", side = 3, line = -2, outer = TRUE, cex = 1.5)

par(mfrow=c(2,2))
plot(hg.lmFi.weight)
mtext("TotHg as a function of weight - Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)


par(mfrow=c(2,2))
plot(hg.lmNF.weight)
mtext("TotHg as a function of weight - Non-Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)
```
##### Comments

#### Height

##### Code

```{r, echo=FALSE, message=FALSE}
hg.form.height = TotHg ~ height

hg.lm.height = lm(hg.form.height, data=dataset)
hg.lmFi.height = lm(hg.form.height, data=dataset.fisherman)
hg.lmNF.height = lm(hg.form.height, data=dataset.non_fisherman)

par(mfrow=c(2,2))
plot(hg.lm.height)
mtext("TotHg as a function of height - Whole population", side = 3, line = -2, outer = TRUE, cex = 1.5)

par(mfrow=c(2,2))
plot(hg.lmFi.height)
mtext("TotHg as a function of height - Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)


par(mfrow=c(2,2))
plot(hg.lmNF.height)
mtext("TotHg as a function of height - Non-Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)
```

##### Comments

#### Fish meal per week

##### Code

```{r, echo=FALSE, message=FALSE}
hg.form.fishmlwk = TotHg ~ fishmlwk

hg.lm.fishmlwk = lm(hg.form.fishmlwk, data=dataset)
hg.lmFi.fishmlwk = lm(hg.form.fishmlwk, data=dataset.fisherman)
hg.lmNF.fishmlwk = lm(hg.form.fishmlwk, data=dataset.non_fisherman)

par(mfrow=c(2,2))
plot(hg.lm.fishmlwk)
mtext("TotHg as a function of fishmlwk - Whole population", side = 3, line = -2, outer = TRUE, cex = 1.5)

par(mfrow=c(2,2))
plot(hg.lmFi.fishmlwk)
mtext("TotHg as a function of fishmlwk - Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)


par(mfrow=c(2,2))
plot(hg.lmNF.fishmlwk)
mtext("TotHg as a function of fishmlwk - Non-Fishermen population", side = 3, line = -2, outer = TRUE, cex = 1.5)
```

##### Comments


### Two groups significantly different

The fishermen have higher levels of mercury in their hair.

Test of the difference between fishermen and non-fishermen 
```{r, echo=FALSE, message=FALSE}
hg.form.fisherman = TotHg ~ fisherman
hg.aov.fisherman = aov(hg.form.fisherman, data= dataset)
summary(hg.aov.fisherman)
```
There is indeed a significant difference between these two groups. What are the differences between the two populations that can explain such observations?

Comments:

- The population of non-fishermen is between 25 and 35 years while the population of fishermen is between 15 and 60 years. As there seem to be a little correlation between age and mercury levels this could affect our other results.

- The variable restime seems difficult to interpret (poor correlation with Hg levels, narrow range of values for non-fishermen).

- The height indicator for non-fishermen is not very precise (it takes only 3 different values: 170cm, 175cm, 180cm).

- There seems to be a correlation between weight and mercury levels. Should we study mercury levels per kg instead? (that may be not very relevant because the mercury levels are the mercury levels in the hair so there have no reason to be linearly correlated with the total mass of the body)

- There is a clear difference in way of life between fishermen and non-fishermen: the first ones eat fish more often than the second. We have to be very careful in interpreting our results because any correlation found between high levels of fish consumption and mercury can reflect the correlation between being a fisherman and having high levels of mercury without meaning that it is fish consumption that causes high levels of Hg. However, among the non-fishermen population there seems to be a clear trend between fish consumption and Hg mercury.

- No clear trend between fishpart and Hg levels, maybe we need to put in relation fishpart and fish consumption.

- Unbalanced design (more fishermen than controls)

```{r, echo=FALSE, message=FALSE}
hg.form.fishpart = TotHg ~ fishpart
hg.aov.fishpart = aov(hg.form.fishpart, data= dataset)
summary(hg.aov.fishpart)
```


## Possible analysis

- Add a correlation coefficient to the scatterplots

- Check homoscedasticity

- Fit a linear model to the data

- Model selection:

  - Compare models using F-tests, AIC, BIC

  - If the number of variables is small enough, could compare all possible models. Usually this is not practical, use automatic procedures: forward selection, backward elimination, stepwise selection

- Adjusted $R^2$, ANOVA

- Look for influential points (studentized residuals, Cook's distance)

- Other diagnostic plots: residuals against predicted values, normal QQ-plot, scale location, residual vs leverage

## Plan 

- VIF to check for multicolinearity between variables + choose which we want to keep 

- stepwise selection on the model with interactions

- fit the model, with the whole population, the fishermen, and the non fishermen

- diagnostic plots

- eventually robust regression 

- conclude for the values of the parameters + some nice plots 

  
### Selection of the model 

```{r, echo=FALSE, message=FALSE}
hg.form.full = TotHg~ age + restime + height + weight + fishmlwk + fishpart 
hg.form.full2 = TotHg~ (age + restime + height + weight + fishpart)^2

vif(lm(hg.form.full2,data=dataset))
```
- The Variance Inflation Factors tends to show there is no case of too high colinearity here. (BUT problem with the fact we have categorical var vs. continuous var ?)

#### Whole population, non-squared model

```{r, echo=FALSE, message=FALSE}
fit <- lm(hg.form.full,data=dataset)
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
plot(step)
```
With this method of stewise selection, it seems that the best model would be : 
TotHg ~ weight + fishmlwk
The Multiple R-squared is only 0.2498: it seems we are unable to explain most of the variability of TotHg between individuals.
However as the p-values for those two parameters are excellents, their influence on TotHg seems well established.

#### Attempt of backward selection

```{r, echo=FALSE, message=FALSE}
hg.form.custom = TotHg ~ age + restime + height + weight + fishmlwk + fishpart
drop1(lm(hg.form.custom, data = dataset), test="F")
# > delete age

hg.form.custom = TotHg ~ restime + height + weight + fishmlwk + fishpart
drop1(lm(hg.form.custom, data = dataset), test="F")
# > delete height

hg.form.custom = TotHg ~ restime + weight + fishmlwk + fishpart
drop1(lm(hg.form.custom, data = dataset), test="F")
# > delete fishpart

hg.form.custom = TotHg ~ restime + weight + fishmlwk
drop1(lm(hg.form.custom, data = dataset), test="F")
# > delete restime

hg.form.custom = TotHg ~ weight + fishmlwk
drop1(lm(hg.form.custom, data = dataset), test="F")
summary(lm(hg.form.custom, data = dataset))
# > seems ok !
# The backward selection gives the same model than stepwise.
```

#### Whole population, squared model

```{r, echo=FALSE, message=FALSE}
fit <- lm(hg.form.full2,data=dataset)
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
```

### Fit of the selected model
#### Code

```{r, echo=FALSE, message=FALSE}
selected.model = TotHg ~ age + restime + height + weight + fishmlwk + fishpart + age:height + age:weight + age:fishpart + restime:weight + restime:fishmlwk + restime:fishpart + height:fishmlwk + height:fishpart + weight:fishpart

hg.lm.selected.model=lm(selected.model, data=dataset)
hg.lmFi.selected.model=lm(selected.model, data=dataset.fisherman)
hg.lmNF.selected.model=lm(selected.model, data=dataset.non_fisherman)

summary(hg.lm.selected.model)
summary(hg.lmFi.selected.model)
summary(hg.lmNF.selected.model)
```

#### Comments 

### Diagnostic plots 
#### Code 

```{r, echo=FALSE, message=FALSE}
plot(hg.lm.selected.model, which=1)
plot(hg.lm.selected.model, which=2)
plot(hg.lm.selected.model, which=3)
plot(hg.lm.selected.model, which=4)
plot(hg.lm.selected.model, which=5)
plot(hg.lm.selected.model, which=6)
```

#### Comments
It seems that there are some influencial points and we sould maybe remove them to analyse the results without them.

### Removing the influencial points 

```{r, echo=FALSE, message=FALSE}
dataset.without.inf=dataset[c(-6,-56,-70,-85),]
dataset.fisherman.without.inf=dataset.fisherman[c(-6,-56,-70,-85),]
dataset.non_fisherman.without.inf=dataset.non_fisherman[c(-6,-56,-70,-85),]

```

### New fit

```{r, echo=FALSE, message=FALSE}
hg.lm.new.selected.model=lm(selected.model, data=dataset.without.inf)
hg.lmFi.new.selected.model=lm(selected.model, data=dataset.fisherman.without.inf)
hg.lmNF.new.selected.model=lm(selected.model, data=dataset.non_fisherman.without.inf)

summary(hg.lm.new.selected.model)
summary(hg.lmFi.new.selected.model)
summary(hg.lmNF.new.selected.model)
```
### New diagnostic plots
```{r, echo=FALSE, message=FALSE}
plot(hg.lm.new.selected.model, which=1)
plot(hg.lm.new.selected.model, which=2)
plot(hg.lm.new.selected.model, which=3)
plot(hg.lm.new.selected.model, which=4)
plot(hg.lm.new.selected.model, which=5)
plot(hg.lm.new.selected.model, which=6)
```

We should maybe use the robust regression because we have now other influencial points.

### Robust regression 
```{r, echo=FALSE, message=FALSE}
hg.rlm.selected.model <- rlm(selected.model, data=dataset)  # robust reg model
#hg.rlmFi.selected.model <- rlm(selected.model, data=dataset.fisherman)
#hg.rlmNF.selected.model <- rlm(selected.model, data=dataset.non_fisherman)

summary(hg.rlm.selected.model)
#summary(hg.rlmFi.selected.model)
#summary(hg.rlmNF.selected.model)
```

Now we have no significant result anymore. Do we remove by hand some more influencial points so that we don't remove everything? 

=> *What is the scientific question we want to answer?*
=> In our model, how do we deal with correlated explanatory variables??

## What I think we should do to answer the "scientific question" (Joseph)
### Model selection
#### Whole population
```{r, echo=FALSE, message=FALSE}
fit <- lm(hg.form.full,data=dataset)
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
plot(step)
```
With this method of stewise selection, it seems that the best model would be : 
TotHg ~ weight + fishmlwk
The Multiple R-squared is only 0.2498: it seems we are unable to explain most of the variability of TotHg between individuals.
However as the p-values for those two parameters are excellents, their influence on TotHg seems well established.

#### Fishermen only

```{r, echo=FALSE, message=FALSE}
fit <- lm(hg.form.full,data=dataset.fisherman)
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
plot(step)
```
For the fishermen population only, with the same method of stewise selection, the same best model is obtained : 
TotHg ~ weight + fishmlwk
The Multiple R-squared is still very low, and the p-values is good only for the weight parameter.
Moreover the weight coefficient is higher than it was in the whole population model. Therefore it seems weight is the dominant explanatory variable among fishermen.
Also we have among fishermen a non-negligible amount of very high values for TotHg > The model do not explain that obviously (maybe we can imagine that a small proportion of the pop are fixing Hg much more / eliminating Hg much slower...)

#### Non-Fishermen only

```{r, echo=FALSE, message=FALSE}
fit <- lm(TotHg~ age + restime + height + weight + fishmlwk, data=dataset.non_fisherman) # hg.form.full
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
plot(step)
```

For the non-fishermen population only, with the same method of stewise selection, a model with weight, fishmlwk and fishpart is obtained. However the values of the parameters for fishpart are absurd, or too strange to be interpreted. If fishpart is removed from the formula, the model obtained is the same as before : 
TotHg ~ weight + fishmlwk
The Multiple R-squared is now very good, and the p-values is good only for the fishmlwk parameter.
Moreover the fishmlwk coefficient is higher than it was in the whole population model. Therefore it seems fishmlwk is the dominant explanatory variable by a very large margin among non-fishermen.



#### Whole population, degree 1

```{r, echo=FALSE, message=FALSE}
hg.form.test <- LogTotHg ~ fisherman:(age + restime + weight + logFishmlwk)
vif(lm(hg.form.test,data=dataset))
fit <- lm(hg.form.test,data=dataset)
step <- stepAIC(fit, direction="both")
step$anova # display results
summary(step)
plot(step)

# hg.rlm.test <- rlm(step$call$formula, data = dataset)
# summary(hg.rlm.test)
# plot(hg.rlm.test)

library(robust)
hg.roblm.test <- lmRob(step$call$formula, data = dataset)
summary(hg.roblm.test)
# plot(hg.roblm.test)
```

I tried three things here:

- I set a log scale on _TotHg_ since it is a value ranging from zero to infinity possibly taking arbitrarily low values, but bottom-bounded by zero: thus it couldn't possibly have a normal distribution for any valuation of the observables.

- I set a (approximate) log scale on _fishmlwk_ since it is a count value ranging from 0 to 21. I followed the idea from the course 5, even if _fishmlwk_ is not a response (*y*) but an observation (*x*).

- I used the robust regression from the _robust_ package in order to obtain the *p-values* for the coefficients of the robust regression, and set a reduced weight to observations classified as outliers by the Cook distance.

I obtained several interesting results:

- The VIFs suggest that no colinearity is to be observed between the observables.

- The right tail of residuals is very reduced and almost fits the normal distribution.

- The left tail of residuals is somewhat expanded, but less than the right tail with non-log scale.

- The residuals **do not appear to be homoscedactically distributed**... Namely, there is an island with small fitted values with a very small variance of residuals...

The final model got by the regression is *`r deparse(step$call$formula)`*.

# References
