---
title: "R Notebook"
output: html_notebook
---

# Imports and loading data

```{r}
# Load here the libraries and packages you need for the rest of the analysis.
library(tidyverse)
library("MASS")
library(GGally)
```


```{r}
DATA_FILE = "fishermen_mercury.csv" # Location of the data csv file
dataset = read.csv(DATA_FILE) # Loading the data
dataset$MeHg <- NULL # We are asked not to use the variable MeHg
dataset$LogTotHg = log(dataset$TotHg)
dataset
```

# Introduction

```{r}
ggplot(dataset, aes(fisherman, TotHg)) + geom_boxplot()
```
```{r}
t.test(x=dataset[which(dataset$fisherman==1),]$TotHg, y=dataset[which(dataset$fisherman==0),]$TotHg, alternative = "greater", var.equal=FALSE)
```
Mercury levels in the hair of fishermen are statistically hgher than in the hair of non-fishermen. How do we explain it?



# Exploratory analysis

## Summarize the dataset
```{r}
summary(dataset)
```


## Pairwise behaviour of explnatory variables
_NB: peut-on remplacer le coefficient de corrélation par l'adjusted $R^2$ ?_



# Analysis: 1-D linear models
_Looking for influential points and testing the hypothesis of the model_
=> Cette partie risque d'être longue et très répétitive. Que garde-t-on dans le rapport ?



# Model
_Use stewise selection to generate the best model_



# Conclusion
_Is our model able to predict the observed means of fishermen and non-fishermen populations observed in the introduction?_
