---
title: 'Project 1: regression 1'
author: "Urvan Christen, Amandine Goffeney, Joseph Vermeil, Lucile Vigu?f©"
date: "March 20, 2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    df_print: paged
fontsize: 12pt
geometry: margin=2.5cm
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[LE,RO]{Christen, Goffeney, Vermeil, Vigu?f©}
df_print: paged
bibliography: bibliography.bib
---

```{r, echo=FALSE, message=FALSE}
# Load here the libraries and packages you need for the rest of the analysis.
library(tidyverse)
library("MASS")
library(GGally)
library(car) # contains the vif() and leveneTest() functions
library(gridExtra)
library(cowplot)
```

```{r, echo=FALSE, message=FALSE}
# Loading data
DATA_FILE = "fishermen_mercury.csv" # Location of the data csv file
dataset = read.csv(DATA_FILE) # Loading the data
dataset$MeHg <- NULL # We are asked not to use the variable MeHg
dataset$LogTotHg <- log(dataset$TotHg)
dataset$logFishmlwk <- log(dataset$fishmlwk + 1)

#Converting categorical variables into factors
dataset$fisherman <- factor(dataset$fisherman)
dataset$fishpart <- factor(dataset$fishpart)
```




# Introduction

Mercury is a metal present in the environment whose harmful effects on human health are well assessed [@park2012human]. In a study led in 2000 [@al2000factors], scientists collected data on total mercury and methyl mercury levels in the hair of 100 fishermen of Kuwait, aged 16 to 58 years, comparing them to those of a control population of 35 non-fishermen, aged 26 to 35 years. The aim of this report is to analyse the factors influencing the levels of mercury in both populations. For the sake of simplicity, we will only focus on total Hg, leaving out methyl mercury, as both variables are strongly correlated.

The dataset gathers information about six numerical variables (age, height, weight, number of fish meals per week and residence time in Kuwait) and two categorical ones (being a fisherman or not, fish consumption habits). There is no additional information about gender as all the participants in the study are males.

A first insight at the data shows that the fishermen population exhibits higher levels of mercury in their hair. The significance of the difference between the means of both distributions is assessed by a _Welch Two Sample t-test_ with the alternative hypothesis that the fishermen population has a average greater level of mercury than the control population (p-value = 7.473e-05).

```{r, echo=FALSE, message=FALSE, fig.width=4, fig.height=2.5, fig.margin=TRUE}
ggplot(dataset, aes(group=fisherman, y=TotHg, fill=fisherman)) + geom_boxplot() + labs(y = "Total Hg concentration in the hair (mg/g)") + scale_x_discrete() + scale_fill_discrete(name = "Population", labels = c("Control group", "Fishermen")) + theme_bw()
```
```{r, include=FALSE}
t.test(x=dataset[which(dataset$fisherman==1),]$TotHg, y=dataset[which(dataset$fisherman==0),]$TotHg, alternative = "greater", var.equal=FALSE)
```


# Exploratory analysis (Amandine)

#### Overview of the data 
Before fitting a model, let's have a look at the data. The following table shows the 8 possible values of fishmlwk according to the 2 possible values of fisherman.
```{r, echo=FALSE, message=FALSE}
table(dataset$fisherman, dataset$fishmlwk)
```
First, let's notice that for some values, we have a very few people, for instance only 2 people eat fish 3 times a week. We see also that the data is really unbalanced, because the fishermen and non fishermen are separated on the variable "fishmlwk": the non fishermen don't eat fish more than 2 times a week, whereas the fishermen don't eat fish less than 3 times a week. 
Besides, we face the same unbalanced pattern for the variables _age_ and _restime_.

```{r, echo=FALSE, message=FALSE}
par(mfrow=c(1,3))
ggplot(dataset, aes(age, TotHg, colour = fisherman)) + geom_point() #+ scale_y_log10() + scale_x_log10() ##Uncomment for log scales
ggplot(dataset, aes(restime, TotHg, colour = fisherman)) + geom_point()
ggplot(dataset, aes(fishmlwk, TotHg, colour = fisherman)) + geom_boxplot()
#grid.arrange(plot1, plot2, plot3, ncol=3, nrow = 1)
```

#### Multicolinearity 

To check whether we have multicolinear variables, we use the VIF function. If all the variables have a result below 5, we keep them. 

```{r ,echo=FALSE, message=FALSE}
hg.form.full = TotHg~ fisherman + age + restime + height + weight + fishmlwk + fishpart 
#hg.form.full2 = TotHg~ (fisherman + age + restime + height + weight + fishpart)^2
#hg.form.full3 = TotHg~ (fisherman + age + restime + height + weight + fishpart)^3

vif(lm(hg.form.full,data=dataset))
#vif(lm(hg.form.full2,data=dataset))
#vif(lm(hg.form.full3,data=dataset))
```
Indeed, we observe that all the variables have a VIF below 2, so we don't eliminate any variable, yet.

# Model selection (Joseph)

Now we are going to use and compare the different methods of model selection to select the more relevant parameters to explain the TotHg variations within the population.

## Stepwise selection

The first attempt is a stepwise selection based on a formula with all the parameters except for "fisherman". Indeed, the preliminary exploration has shown that the "fisherman" variable have considerable impact on many of the other variables. Therefore to understand what is the reason behind the TotHg difference between the 2 groups, it seems natural to start with a model without the "fisherman" variable.

```{r, echo=FALSE, message=FALSE}
hg.form.step1 = TotHg ~ age + restime + height + weight + fishmlwk + fishpart
model1 <- stepAIC(lm(hg.form.step1,data=dataset), direction="both")
summary(model1)
```

With stepwise selection, the selected model is : TotHg ~ weight + fishmlwk. The intercept and the two coefficients are very significant. Moreover the signs of the coefficients are not absurd: while it is not really intuitive that the weight coefficient should be positive or negative, the coefficient of fishmlwk had to be positive, and it's the case here.

Now to determine possible differences between the fishermen and non-fishermen, a model based on the interactions between the "fisherman" variable and all the others is proposed. This could show if a certain variable is very relevant concerning fishermen but less when it comes to non-fishermen, for instance.

```{r, echo=FALSE, message=FALSE}
hg.form.step2 = TotHg ~ fisherman:(age + restime + height + weight + fishmlwk + fishpart)
model2 <- stepAIC(lm(hg.form.step2,data=dataset), direction="both")
summary(model2)
```

The best model is now : TotHg ~ fisherman:weight + fisherman:fishmlwk, ie the interactions between fisherman and the variables of the previous model. While the coefficients for fisherman0:weight and fisherman1:weight stays very significant, fisherman0:fishmlwk and fisherman1:fishmlwk are barely significant (but their p-value is very close to 0.05) ; this comes from the fact that within each of the subpopulation {fishermen} and {non-fishermen} the range in which fishmlwk varies is very reduced compared to the whole population (see the Exploratory analysis). Moreover, when looking at the values of the coefficients, it appears that for the non-fishermen, the number of fish meal per week has a very preponderant effect, while among fishermen, the contribution of weight and fish meal per week are almost even.

In order to check the dependency of the selected model to the selection technique, we have tried backward and forward selection and obtained similar models.



```{r include=FALSE}
hg.form.bwd = TotHg ~ fisherman:(age + restime + height + weight + fishmlwk + fishpart)
drop1(lm(hg.form.bwd, data = dataset), test="F")
#> Delete fisherman:fishpart

hg.form.bwd = TotHg ~ fisherman:(age + restime + height + weight + fishmlwk)
drop1(lm(hg.form.bwd, data = dataset), test="F")
#> Delete fisherman:age

hg.form.bwd = TotHg ~ fisherman:(restime + height + weight + fishmlwk)
drop1(lm(hg.form.bwd, data = dataset), test="F")
#> Delete fisherman:restime

hg.form.bwd = TotHg ~ fisherman:(height + weight + fishmlwk)
drop1(lm(hg.form.bwd, data = dataset), test="F")
#> Delete fisherman:height

hg.form.bwd = TotHg ~ fisherman:(weight + fishmlwk)
drop1(lm(hg.form.bwd, data = dataset), test="F")
#> All parameters are significant (p-value < 0.05)
```


```{r include=FALSE}
hg.form.fwd = TotHg ~ 1
add1(lm(hg.form.fwd, data=dataset), test="F", scope = ~ fisherman:(age + restime + height + weight + fishmlwk + fishpart))
#> Add fisherman:weight

hg.form.fwd = TotHg ~ fisherman:weight
add1(lm(hg.form.fwd, data=dataset), test="F", scope = ~ fisherman:(age + restime + height + weight + fishmlwk + fishpart))
#> Add fisherman:fishmlwk

hg.form.fwd = TotHg ~ fisherman:weight + fisherman:fishmlwk
add1(lm(hg.form.fwd, data=dataset), test="F", scope = ~ fisherman:(age + restime + height + weight + fishmlwk + fishpart))
#> No more significant (p-value < 0.05) parameter to add
```


# Results and discussion (Urvan)

Fit model (* fishermen) (I picked the model given by stepwise selection)

```{r, echo=FALSE, message=FALSE}
hg.form.selected=TotHg ~ fisherman + weight + fishmlwk + fisherman:weight + fisherman:fishmlwk
hg.lm.selected.model=lm(hg.form.selected, data=dataset)
summary(hg.lm.selected.model)
```


Commenter les résultats, valeurs des coefficients, font-elles du sens, p-values

Les diagnostic plots pour weight + dire que c'est pareil pour tout le monde

```{r, echo=FALSE, message=FALSE}
plot(hg.lm.selected.model, which=1) #residuals vs fitted
plot(hg.lm.selected.model, which=2) #QQ plot
plot(hg.lm.selected.model, which=3) #scale location
plot(hg.lm.selected.model, which=4) #cook's distance
plot(hg.lm.selected.model, which=5) #residuals vs leverage 
plot(hg.lm.selected.model, which=6) #cook's distance vs leverage
```


# Conclusion (Lucile)

3 lignes

Les principales variables explicatives sont:

- La consommation de poisson => logique
- Le poids => plus étonnant mais on sait que les cellules graisseuses stockent les toxiques

# References